\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-28 move_schedules.tex
Cleaned up reduction section

 The main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagram is encoded in $R$, and the generating cycles in the columns of $V$. 
 Given a filtration $(K, f)$ with $m$ simplices, one assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. 

@@

% The reduction algorithm computes the persistent homology from $D$ for all dimensions up to $\mathrm{dim}(K) - 1$.
%  Alternatively, a single dimension $p \leq \mathrm{dim}(K) - 1$ can be computed, in which case one reduces a pair of sub-matrices $D_p$ and $ D_{p+1}$ of $D$,  where $D_p$ has dimension $d_{p-1} \times d_p$ and $D_{p+1}$ is $d_{p} \times d_{p+1}$. 
%In some situations the description of the ensuing algorithms is simpler when considering all dimensions simultaneously, while in others it is more succinct to consider dimension-specific matrices. 
%In what follows $D$, $R$, and $V$ will always denote square $(m \times m)$ matrices, though we note that all algorithms discussed generalize to both cases. 

@@

and whose filtration order, read as inclusions from left to right, is given as $K = (u, v, w, a, b, c)$. The reduction algorithm begins by setting $R = D$ and $V = I$, followed by left-to-right column operations until invariants I3 is satisfied.
If the $i$-th column of $R$ is nonzero, then $\mathrm{low}_R(i)$ 
is the row index of its lowest nonzero entry.
Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 for simplicity, the reduction proceeds as follows:

@@

Nonetheless, many variations and optimizations to Algorithm~\ref{alg:reduce} have been proposed over the past decade, see~\cite{bauer2017phat, bauer2021ripser, chen2011persistent} for an overview. Despite these variations in computation, the resulting decomposition must satisfy the following invariants:
  \vspace*{0.8em}
  \textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $K$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 \vspace*{1em}
We call the decomposition satisfying these three invariants \emph{valid}. 
%%\begin{proposition}[Persistent Pairing~\cite{zomorodian2005computing}]
%%	Let $R = D V$, where $R$ is in reduced form and $V$ is full-rank upper triangular (invariants (1) and (2)  hold). Then the pairing $(\sigma_i, \sigma_j)$ defined by $i = \mathrm{low}_R(j)$ is a \emph{creator/destroyer (persistent) pair}. While the matrices $R,V$ satisfying (1) and (2) are not unique in general,  the persistent pairs $(i,j)$ are unique and define the persistence diagrams of $K$.
%%\end{proposition}
Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}. The persistence diagram of the corresponding filtration $K$ may be read off from lowest entries in $R$, once $R$ is in reduced form.

@@

These rules prescribe certain column and row operations to apply to a given matrix decomposition either before, during, or after each individual transposition to ensure it is a valid decomposition.   

@@

%Lesnick and Wright~\cite{lesnick2015interactive} introduced an elegant reparameterization that not only fully characterizes $\mathcal{B}(M)$, but also yields an efficient method for rendering $\mathcal{B}_L(M)$ in real time. 
Utilizing the equivalence between the rank and fibered barcode invariants, Lesnick and Wright~\cite{lesnick2015interactive} developed an elegant way of computing $\mathcal{B}(M)$ via a reparameterization using standard point-line duality. This clever technique effectively reduces the fibered barcode computation to a sequence of 1-D barcode computations at ``template points'' lying within the 2-cells of a particular planar subdivision $\mathcal{A}(M)$ of the half-plane $[0, \infty) \times \mathbb{R}$. This particular subdivision is induced by the arrangement of ``critical lines'' derived by the bigraded Betti numbers $\beta(M)$ of $M$.
As the barcode of one template point $\mathcal{T}_e$ at the 2-cell $e \in \mathcal{A}(M)$ may be computed efficiently by re-using information from an adjacent template point $\mathcal{T}_{e'}$,~\cite{lesnick2015interactive} observed that computing the barcodes of all such template points (and thus, $\mathcal{B}(M)$) may be reduced to ordering the 2-cells in $\mathcal{A}(M)$ along a Eulerian path traversing the dual graph of $\mathcal{A}(M)$. 
The full algorithm is out of scope for this effort; we include supplementary details for the curious reader in the appendix~\ref{app:2d_pers}. 
\\
\\
\noindent 
\textbf{Example}: Consider a small set of noisy points distributed around $S^1$ containing a few strong outliers, as shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose various invariants are shown in the middle figure. The gray areas indicate homology with positive dimension---the lighter gray area $\mathrm{dim}_1(M) = 1$ indicates a persistent loop was detected.
On the right side, dual space is shown: the black lines are the critical lines that form $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph of $\mathcal{A}(M)$, the rainbow lines overlaying the dashed-lines form the Eulerian path, and the orange barycentric points along the 2-cells of $\mathcal{A}(M)$ represent where the barcodes templates $\mathcal{T}_e$ are parameterized. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the dimension function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
%The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\

>>>>

 Given a filtration $(K, f)$ with $m$ simplices,  the main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagram is encoded in $R$ and the generating cycles in the columns of $V$.
 To begin the reduction, one starts by assembling the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. Setting $V = I$ and $R = D$, one proceeds by performing elementary left-to-right column operations on $V$ and $R$ until the following invariants are satisfied:
\textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $K$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 	\end{enumerate} 
 \vspace*{1em}
 where $\mathrm{low}_R(i)$ denotes the largest row index of a non-zero entry in column $i$ of $R$. 
We call the decomposition satisfying these three invariants \emph{valid}. Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}. The persistence diagram of the corresponding filtration $K$ may be read off from lowest entries in $R$, once $R$ is in reduced form.

@@

@@

and whose filtration order is given as $K = (u, v, w, a, b, c)$. 
%The reduction algorithm begins by setting $R = D$ and $V = I$, followed by left-to-right column operations until invariants I3 is satisfied.
%If the $i$-th column of $R$ is nonzero, then $\mathrm{low}_R(i)$  is the row index of its lowest nonzero entry.
Using $\mathbb{Z}_2$ coefficients for simplicity, the reduction proceeds to compute $H_1(K)$ as follows:

@@

Despite this high algorithmic complexity, the number of column operations has been observed to be super-linear in practice, due impart to the high sparsity and structure of $D$. 
Moreover, many variations and optimizations to Algorithm~\ref{alg:reduce} have been proposed over the past decade, see~\cite{bauer2017phat, bauer2021ripser, chen2011persistent} for an overview. 

@@

These rules prescribe certain column and row operations to apply to a given matrix decomposition either before, during, or after each transposition to ensure the decomposition is valid.   

@@

%Lesnick and Wright~\cite{lesnick2015interactive} introduced an elegant reparameterization that not only fully characterizes $\mathcal{B}(M)$, but also yields an efficient method for rendering $\mathcal{B}_L(M)$ in real time. 
Utilizing the equivalence between the rank and fibered barcode invariants, Lesnick and Wright~\cite{lesnick2015interactive} developed an elegant way of computing $\mathcal{B}(M)$ via a reparameterization using standard point-line duality. This clever technique effectively reduces the fibered barcode computation to a sequence of 1-D barcode computations at ``template points'' lying within the 2-cells of a particular planar subdivision $\mathcal{A}(M)$ of the half-plane $[0, \infty) \times \mathbb{R}$. This particular subdivision is induced by the arrangement of ``critical lines'' derived by the bigraded Betti numbers $\beta(M)$ of $M$.
As the barcode of one template point $\mathcal{T}_e$ at the 2-cell $e \in \mathcal{A}(M)$ may be computed efficiently by re-using information from an adjacent template point $\mathcal{T}_{e'}$,~\cite{lesnick2015interactive} observed that computing the barcodes of all such template points (and thus, $\mathcal{B}(M)$) may be reduced to ordering the 2-cells in $\mathcal{A}(M)$ along a Eulerian path traversing the dual graph of $\mathcal{A}(M)$. 
The full algorithm is out of scope for this effort; we include supplementary details for the curious reader in the appendix~\ref{app:2d_pers}. 
\\
\\
\noindent 
\textbf{Example 4.1}: Consider a small set of noisy points distributed around $S^1$ containing a few strong outliers, as shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose various invariants are shown in the middle figure. The gray areas indicate homology with positive dimension---the lighter gray area $\mathrm{dim}_1(M) = 1$ indicates a persistent loop was detected.
On the right side, dual space is shown: the black lines are the critical lines that form $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph of $\mathcal{A}(M)$, the rainbow lines overlaying the dashed-lines form the Eulerian path, and the orange barycentric points along the 2-cells of $\mathcal{A}(M)$ represent where the barcodes templates $\mathcal{T}_e$ are parameterized. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the dimension function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
%The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\
